---
title: "Spatial data in R"
author: "Christian König & Damaris Zurell"
output: rmarkdown::github_document
---

**This practical exercise is a part of the workshop [Big Data Ecology](https://github.com/ChrKoenig/Big_Data_Ecology)**

---
```{r global_options, include = F}
knitr::opts_chunk$set(fig.path='../figs/') # Save output figures for rendering .md document
```
## Introduction
Here, we focus on spatial data, a central data type in ecology. Following the content of the lecture, we will work with vector and raster data and perform a number of operations based on their attributes, geometries and spatial relationships.

We will use the following packages:

```{r, message=FALSE}
library(sf)
library(raster)
library(rnaturalearth)
library(dplyr)
library(ggplot2)
```

If you haven't installed them, please do so with the following command: 

```{r, eval = F}
install.packages(c("sf", "raster", "rnaturalearth", "dplyr", "ggplot2"))
```

## Vector data
Vector data typically represent discrete objects with clear boundaries such as, e.g., individual trees, roads, or countries. In addition to their spatial definition (geometry), vector data may carry ancillary data on the spatial features, e.g. the species of the tree, the type of the road, or name of the country. These additional information are called attributes. 

### Vector data types
There are three fundamental types of vector data:

* **Points** occur at discrete locations and are represented by a coordinate pair (x,y). 
* **Lines** describe linear features and are defined by at least two coordinate pairs (x,y), the end points of the line. A line can also consist of several line segments.
* **Polygons** describe two-dimensional features in the landscapes and define a bounded area, enclosed by lines. Thus, a polygon needs to consist of at least three coordinate pairs (x,y).

The `sf` package provides a set of vector data classes and methods (*simple features*) that follow the *tidy* data model we already met with the `dplyr` package. In fact, the *simple feature* standard is language-independent and has been [adopted by many geospatial analysis and database platforms](https://en.wikipedia.org/wiki/Simple_Features#Implementations). You can think of a *simple feature* as a normal data frame where each row corresponds to one spatial entity, with the coordinates/geometries of these entities being stored in a special column. Additional attributes can be simply added as new columns. This design enables the use of tidy data manipulation tools for *simple features* without breaking the link between the attribute and geometry data.  

### Point features
Let's create a set of 20 points with random longitude and latitude values:
```{r}
set.seed(1)
coords = data.frame(
  longitude = runif(n = 20, min = -180, max = 180),
  latitude = runif(n = 20, min = -90, max = 90)
)
plot(coords)
```

Now we use the `sf`package to convert this coordinate data frame into a collection of simple point features.

```{r}
points_sf = sf::st_as_sf(coords, coords = c("longitude", "latitude"), crs = 4326)  
points_sf
```

The `print` method outputs an informative summary of our new `sf` object: It's a simple feature collection of 20 point features. These point features are defined in two dimensions (XY) and have no additional attributes. Also note that the EPSG code provided to the `CRS` argument of the `st_as_sf()` function is sufficient to fully define the coordinate reference system (CRS) of our spatial object.

Next, we use the very same manipulation tools as for other tidy data to add an *attribute* column to `points_sf`. Specifically, we use the `mutate()` verb known from `dplyr` for this task:

```{r}
points_sf = points_sf %>% 
  mutate(name = sample(LETTERS[1:5], 20, replace = T)) # Add random letters as new attribute 'name'
points_sf
```

### Line and Polygon features
The structure of simple line and polygon features is analogous to simple point features, the only difference being that the geometry column now holds multiple coordinates to define the shape of the features. We'll confirm this by aggregating `points_sf` to one line feature per name.

```{r}
lines_sf = points_sf %>% 
  group_by(name) %>%     # group by name
  filter(n() > 1) %>%    # lines need more than 1 point
  summarise(do_union = F) %>%  # Don't union geometries
  sf::st_cast("LINESTRING")  # Create one line feature per group
lines_sf
```

Finally, let's plot our spatial objects. The `sf` package implements a custom geom class that works with the `ggplot2` package:

```{r}
ggplot() +
  geom_sf(data = lines_sf, aes(color = name)) + # Plot lines, color by group
  geom_sf(data = points_sf)   # Plot points in standard black
```

## Raster data
Raster data are used to represent grids of values across a continuous area, e.g. a digital elevation model, NDVI satellite measurements, or interpolated climate variables. We will use the `raster` package to represent and analyse raster data in `R`. The package contains different data classes, most importantly `RasterLayer`, `RasterStack` and `RasterBrick`. `RasterLayers` contain only a single layer of values while `RasterStacks` and `RasterBricks` contain multiple layers (from separate files or from a single multi-layer file, respectively).

### RasterLayers, RasterStacks and RasterBricks
The function `raster()` can be used to create `RasterLayer` objects. We'll create a continuous global grid of 10x10-degree raster cells. For illustration purposes, we set the CRS with a proj4 string this time.

```{r}
r1 = raster::raster(ncol=36, nrow=18, xmx=180, xmn=-180, ymx=90,  ymn=-90)
raster::crs(r1) = "+proj=longlat +datum=WGS84 +no_defs" # set CRS with Proj4 string
r1
```

`RasterLayers` can only possess one attribute/value per cell. We use the `values()` function to access the attributes of our empty raster object and assign some random values.

```{r}
values(r1) = rnorm(ncell(r1)) # assign random values
plot(r1)   # plot raster
```

`RasterStack` and `RasterBrick` objects can be created with the `stack()` and `brick()` function, respectively. Note that this only works when all rasters have the same spatial extent and resolution.

```{r, results='hide'}
r2 = r1 # Copy r1 into new object

raster::stack(r1,r2) # Create RasterStack 
raster::brick(r1,r2) # Create RasterBrick
```

### Importing and downloading data
The `raster()`, `stack()`, and `brick()` functions can not only be used for creating `Raster*` objects from scratch, but also for reading raster files from disk. To do this, simply provide one or multiple file paths instead of the names of the `R` objects as function arguments.

Additionally, the `raster` package offers the interesting feature of downloading data directly from a number of standard repositories with the `getData()` function. For more information, see the help pages with `?getData`.

```{r}
# Download RasterStack of global monthly minimum temperatures (°C*10) from WorldClim
tmin = raster::getData("worldclim", var="tmin", res=10, download = T, path = "../data/")
plot(tmin)
```

## Manipulating vector and raster data
In the following section, we will look at some common operations on spatial data. 

### Attribute operations
As we've seen above, we can manipulate the attribute data of `sf` objects in the same way as non-spatial tidy data by using `dplyr`-syntax, e.g. the `filter()`, `mutate()`, `select()`, `summarise()`, and `arrange()` functions. The special geometry column of `sf` objects is *sticky* throughout attribute data operations, i.e. it is always added back to the result. If we want to explicitly remove the spatial context of the data, e.g. to speed up computations, we can use the `st_drop_geometry()` function and convert our `sf` object to an ordinary `data.frame`.

```{r}
points_sf %>% 
  sf::st_drop_geometry() %>% 
  class()
```

Attribute data of raster objects behave similar (but not identical!) to matrices in `R`. We can do arithmetic operations (addition, subtraction, multiplication, division), produce layer-wise summaries with the `cellStats()` function, or calculate cell-wise summaries with the standard `min()`, `max()`, or `mean()` functions. Both single- and multi-layer `rasters` can be subset using familiar `[]`-syntax`.

```{r, results='hide'}
r1 + r1              # Arithmetic
raster::cellStats(r1, min)   # Minimum value per layer
r1_stack = min(stack(r1, r1/2)) # Minimum value per cell 
r1_stack[[1]]        # Subset by layer index
r1[7,13]             # Subset by row/column index
```

### Spatial data and geometry operations
Instead of their attributes, we can also use the shape and location of spatial objects to modify either their attribute data (spatial data operations) or their geometries (geometry operations). A complete coverage of all spatial operations is far beyond the scope of this workshop, so we'll just have a look at a few examples.

First, we'll spatially join country-level information to our `points_sf` layer based on the points' coordinates, i.e. in which country they fall. We use the `ne_countries()` function from the `rnaturalearth` package to download a *simple feature* polygon layer containing a number of country-level attribute data, and then use the `st_join()` function to merge the country name to our point layer:

```{r}
world_sf = rnaturalearth::ne_countries(returnclass = "sf") # Download country layer using rnaturalearth package
st_join(points_sf, world_sf) %>%            # Join data based on location, not attribute columns
  dplyr::select(point_name = name.x, country_name = name.y) # select and rename results columns
```

All spatial operations in the `sf` package have the prefix `st` for 'spatial type'. A full list of `st`-functions (and everything else...) in `sf` can be found [here](https://r-spatial.github.io/sf/reference/index.html).

Let's also look at an example of a spatial operation on raster data. Processing raster data is very resource-intensive at high spatial resolutions and we might find ourselves in a situation where we need to decrease the resolution of a `Raster*` object to speed up computations. In this case we could use the `aggregate()` function and coarsen the raster by some factor

```{r}
tmin_coarse = raster::aggregate(tmin, fact = 50, na.rm = T)
```

or we can `resample()` its resolution to that of another raster object. This is very convenient if we want to `stack()` raster layers of different resolutions.

```{r}
tmin_resampled = raster::resample(tmin, r1)
```

### Raster-Vector operations
We can extract raster values for a set of point, line or polygon features with the `extract()` function. 

```{r, results='hide'}
raster::extract(r1, points_sf) # Extract values of r1 at coordinates of points_sf
```

For lines and polygons, we can additionally supply a summary function to the `fun` argument of `extract()`, e.g. to calculate the mean values per polygon.

One last important operation is the cropping and masking of `Raster*` objects. Both of these operations are a form of geometric subsetting, but the `crop()` function uses the *spatial extent* whereas the `mask()` function uses the *geometry* of another spatial object.

```{r, warning = FALSE}
sweden = world_sf %>% filter(name == "Sweden")
plot(crop(tmin[[1]], sweden)) # Subset by extent
plot(mask(tmin[[1]], sweden)) # Subset by geometry, extent unchanged
```

## Exercise

* Create a `RasterStack` from `r1` and `tmin` (Tip: you need to match both the extent and the resolution of both layers)
* Which cell in `tmin` has the lowest mean value across the year (Tip: First calculate the cell-wise mean and then the layer-wise minimum) 
* Add a new attribute to `world_sf` containing the average minimum temperature in January of each country (Tip: Subset `tmin` to the correct layer and think about coarsening its resolution, run the extraction within `mutate()` to create a new attribute). What are the only two countries with a value below -30 C°?

## Some final notes
Spatial data are among the computationally most challenging data types in ecology. This is exacerbated by the advent of very high-resolution remote sensing data with resolutions of up to 10 m. Understanding how to handle them effectively can mean the difference between seconds and hours of compute time. 